---
title: "sea-gtfs-proj-data-cleaning"
author: "Peter Silverstein"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
# Loading Libraries

# General Use
library(tidyverse)
library(ggplot2)
library(here)
library(keyring)

# PostgreSQL
library(DBI)
library(RPostgres)

# GIS and Mapping
library(sf)
library(tmap)

# Census
library(tidycensus)
```

# Main Dataset

## Importing Data From SQL

```{r}
con <- dbConnect(RPostgres::Postgres(),
                 dbname = "sea-gtfs-data",
                 host = "localhost",
                 port = 5432,
                 user = "postgres",
                 password = "Parkour")

gtfs_realtime <- dbReadTable(con, "sea_gtfs_data")
gtfs_realtime <- tibble(gtfs_realtime)
```

## Cleaning Data/Verifying Types/Removing Duplicates

```{r}
# NEED TO FIGURE OUT TIMES OVER 24H!
gtfs_main <- gtfs_realtime %>%
  filter(projection == FALSE) %>%
  distinct(trip_id, stop_id, arrival_datetime, .keep_all = TRUE) %>%
  select(trip_id, route_id, stop_id, arrival_datetime, departure_datetime, pull_datetime) %>%
  mutate(trip_id = as.factor(trip_id),
         route_id = as.factor(route_id),
         stop_id = as.factor(stop_id)) %>%
  mutate(arrival_datetime = with_tz(arrival_datetime, "America/Los_Angeles"),
         departure_datetime = with_tz(departure_datetime, "America/Los_Angeles"),
         pull_datetime = with_tz(pull_datetime, "America/Los_Angeles"))

print(paste("Number of rows reduced from", nrow(gtfs_realtime), "to", nrow(gtfs_main)))
```

# Stop Times Dataset

## Importing Stop Times

```{r}
stop_times <- read_csv("Other Data/gtfs-static-files/stop_times.txt") %>%
  select(trip_id, arrival_time, departure_time, stop_id, stop_sequence, shape_dist_traveled) %>%
  mutate(trip_id = as.factor(trip_id),
         stop_id = as.factor(stop_id)) %>%
  rename(sched_arrival_time = arrival_time,
         sched_departure_time = departure_time) %>%
  distinct(trip_id, stop_id, .keep_all = TRUE)
```

## Joining to Main Dataset

```{r}
gtfs_main_withstoptimes <- gtfs_main %>%
  left_join(stop_times, by = c("trip_id" = "trip_id",
                              "stop_id" = "stop_id"))
```

## Calculating Delay

```{r}
gtfs_main_withdelays <- gtfs_main_withstoptimes %>%
  mutate(actual_arrival_time = hms::as_hms(format(with_tz(arrival_datetime, "America/Los_Angeles"), "%H:%M:%S")),
         actual_departure_time = hms::as_hms(format(with_tz(departure_datetime, "America/Los_Angeles"), "%H:%M:%S")),
         date = as.POSIXct(format(with_tz(arrival_datetime, "America/Los_Angeles"), "%Y-%m-%d"))) %>%
  mutate(arrival_delay = as.numeric(actual_arrival_time - sched_arrival_time)) %>%
  select(date, 
         trip_id, 
         route_id, 
         stop_id, 
         sched_arrival_time, 
         sched_departure_time,
         actual_arrival_time,
         actual_departure_time,
         arrival_delay,
         stop_sequence,
         shape_dist_traveled,
         pull_datetime)
```

## Checking if Join worked, whether all rows have values for all fields

```{r}
test_df <- gtfs_main_withdelays %>%
  filter(is.na(date))

test_df

# Notes
  # 7626 rows have no date/arrival time
    # could this mean a skipped trip?
    # could not discern a pattern to the missing data, but I note that this pattern would appear in cases when stops        were skipped or routes were cancelled
  # 3915 rows have no route_id
    # could probably look up route_id based on trip_id
  # Stop ID 1746 has no stop_times join
    # I cannot discern why this is, exactly, but it's a very small portion of the sample
```

# Traffic by Time Bin

## Loading the dataset

```{r}
congestion_temporal <- read_csv("Other Data/Traffic_Count_Studies_by_Hour_Bins-2.csv")
```

## Table w/ average traffic volume by day/hour

```{r}
congestion_dayhour <- congestion_temporal %>%
  mutate(datetime = as.POSIXct(ADD_DTTM, 
                               format = "%m/%d/%Y %I:%M:%S %p", 
                               tz = "America/Los_Angeles")) %>%
  filter(datetime > as.POSIXct("01-01-2015 00:00:00", 
                               format = "%m-%d-%Y %H:%M:%S",
                               tz = "America/Los_Angeles")) %>%
  group_by(WEEKDAY) %>%
  summarize(HR01 = mean(HR01_TOTAL),
            HR02 = mean(HR02_TOTAL),
            HR03 = mean(HR03_TOTAL),
            HR04 = mean(HR04_TOTAL),
            HR05 = mean(HR05_TOTAL),
            HR06 = mean(HR06_TOTAL),
            HR07 = mean(HR07_TOTAL),
            HR08 = mean(HR08_TOTAL),
            HR09 = mean(HR09_TOTAL),
            HR10 = mean(HR10_TOTAL),
            HR11 = mean(HR11_TOTAL),
            HR12 = mean(HR12_TOTAL),
            HR13 = mean(HR13_TOTAL),
            HR14 = mean(HR14_TOTAL),
            HR15 = mean(HR15_TOTAL),
            HR16 = mean(HR16_TOTAL),
            HR17 = mean(HR17_TOTAL),
            HR18 = mean(HR18_TOTAL),
            HR19 = mean(HR19_TOTAL),
            HR20 = mean(HR20_TOTAL),
            HR21 = mean(HR21_TOTAL),
            HR22 = mean(HR22_TOTAL),
            HR23 = mean(HR23_TOTAL),
            HR24 = mean(HR24_TOTAL)) %>%
  mutate(WEEKDAY_NAME = case_when(
    WEEKDAY == 1 ~ "Monday",
    WEEKDAY == 2 ~ "Tuesday",
    WEEKDAY == 3 ~ "Wednesday",
    WEEKDAY == 4 ~ "Thursday",
    WEEKDAY == 5 ~ "Friday",
    WEEKDAY == 6 ~ "Saturday",
    WEEKDAY == 7 ~ "Sunday")) %>%
  mutate(AVG_VOL = select(., HR01:HR24) %>% rowMeans(na.rm = TRUE)) %>%
  relocate(WEEKDAY_NAME, .after = WEEKDAY) %>%
  relocate(AVG_VOL, .after = WEEKDAY_NAME)

congestion_dh_longer <- congestion_dayhour %>%
  select(WEEKDAY_NAME, HR01:HR24) %>%
  pivot_longer(cols = -WEEKDAY_NAME,
               names_to = "HOUR",
               values_to = "avg_traffic_dayhour") %>%
  mutate(HOUR = as.numeric(gsub("[^0-9]","", HOUR)))
```

## Joining Day/Daypart Congestion

```{r}
# MAKE SURE TO REMOVE -1 VALUES
gtfs_main_withcongestion <- gtfs_main_withdelays %>%
  mutate(WEEKDAY = weekdays(date),
         HR = as.numeric(sched_arrival_time) %/% 3600) %>%
  left_join(select(.data = congestion_dayhour, WEEKDAY_NAME, AVG_VOL), 
            by = c("WEEKDAY" = "WEEKDAY_NAME")) %>%
  left_join(congestion_dh_longer,
            by = c("WEEKDAY" = "WEEKDAY_NAME",
                   "HR" = "HOUR")) %>%
    rename("weekday" = "WEEKDAY",
         "hr" = "HR",
         "avg_traffic_day" = "AVG_VOL")
```

# Route Lines/Stop Points Shapefiles

## Loading Stop Shapefile

```{r}
kingcountymetro_stops <- st_read(here("Other Data",
                                      "KCMetro_Transit_Stops",
                                      "Transit_Stops_for_King_County_Metro___transitstop_point.shp"))
```

```{r}
# Select Columns; NOTE projected CRS is NAD83
kcm_stops <- kingcountymetro_stops %>%
  select(STOP_ID, STOP_STATU, XCOORD, YCOORD, geometry) %>%
  filter(STOP_STATU == "ACT") %>%
  select(!STOP_STATU) %>%
  mutate(STOP_ID = as.factor(STOP_ID)) %>%
  distinct()
```

## Loading Route Shapefile

```{r}
kingcountymetro_routes <- st_read(here("Other Data",
                                      "KCMetro_Transit_Lines",
                                      "Transit_Routes_for_King_County_Metro___transitroute_line.shp"))
```

```{r}
# Select Columns; NOTE projected CRS is NAD83
kcm_routes <- kingcountymetro_routes %>%
  select(IN_SERVICE, ROUTE_ID, LOCAL_EXPR, ROUTE_NUM, SHAPE_Leng, geometry) %>%
  filter(IN_SERVICE == "Y") %>%
  select(!IN_SERVICE) %>%
  mutate(ROUTE_ID = as.factor(ROUTE_ID)) %>%
  distinct(ROUTE_ID, .keep_all = TRUE)
```

## Joining

```{r}
gtfs_main_withstopshapes <- gtfs_main_withdelays %>%
  left_join(kcm_stops, by = c("stop_id" = "STOP_ID"))

gtfs_main_withrouteshapes <- gtfs_main_withdelays %>%
  left_join(kcm_routes, by = c("route_id" = "ROUTE_ID"))
```

```{r}
test_df <- gtfs_main_withrouteshapes %>%
  filter(is.na(ROUTE_NUM))

test_df

# STOPS: Appears as though the only rows with issues here are ones with stop_id 1746
# ROUTES: Appears as though any NAs from join also have NA for route_id (i.e., no extra)
```

# Traffic Flow Counts by Arterial

## Loading Data

```{r}
congestion_spatial_initial <- st_read(here("Other Data",
                                   "2018_Traffic_Flow_Counts-shp",
                                   "2018_Traffic_Flow_Counts.shp"))
```

```{r}
# Select Columns; NOTE projected CRS is NAD83
congestion_spatial <- congestion_spatial_initial %>%
  select(STNAME_ORD, DOWNTOWN, AWDT, ADT, DATAQUALIT, FLAGS, SHAPE_Leng, geometry)
```

# Census Bureau 5-Year Estimates (2019-2023)
```{r}
# Using keyring package to keep my API key hidden
tidycensus_api_key <- key_get(service = "tidycensus_API", username = "my_tidycensus")
census_api_key(tidycensus_api_key)

ACSlist <- load_variables(2022, "acs5", cache = TRUE)
```

## Loading ACS Data from Tidycensus

```{r}
# Projection is NAD83(!!)
kingcounty_acs_blocks <- get_acs(state = "WA",
                                 county = "King",
                                 geography = "block group",
                                 variables = c(total_population = "B01003_001",
                                               median_income = "B19013_001",          # also have poverty level if needed
                                               race_basetotal = "B02001_001",
                                               race_white = "B02001_002",
                                               race_black = "B02001_003",
                                               race_nativeamerican = "B02001_004",
                                               race_asian = "B02001_005",
                                               race_nativehawaiianpi = "B02001_006",
                                               hisp_basetotal = "B03003_001",
                                               hisp_hispanic = "B03003_003",
                                               transp_basetotal = "B08134_001",
                                               tt_less10 = "B08134_002",
                                               tt_10to14 = "B08134_003",
                                               tt_15to19 = "B08134_004",
                                               tt_20to24 = "B08134_005",
                                               tt_25to29 = "B08134_006",
                                               tt_30to34 = "B08134_007",
                                               tt_35to44 = "B08134_008",
                                               tt_45to59 = "B08134_009",
                                               tt_60plus = "B08134_010",
                                               transp_mthd_carbusvan = "B08134_011",
                                               transp_mthd_public = "B08134_061",     # can break this down further by mthd
                                               transp_mthd_public_bus = "B08134_071", # can break this down further by tt
                                               agg_tt_overall = "B08136_001",
                                               agg_tt_public = "B08136_007",
                                               agg_tt_public_bus = "B08136_008",
                                               public_assistance_basetotal = "B19123_001",
                                               public_assistance_yes = "B19123_002",
                                               tenure_basetotal = "B25003_001",
                                               tenure_ownerocc = "B25003_002",
                                               tenure_renterocc = "B25003_003",
                                               units_instructure_basetotal = "B25024_001",
                                               units_instructure_1det = "B25024_002",
                                               units_instructure_1att = "B25024_003",
                                               units_instructure_2 = "B25024_004",
                                               units_instructure_3to4 = "B25024_005",
                                               units_instructure_5to9 = "B25024_006",
                                               units_instructure_10to19 = "B25024_007",
                                               units_instructure_20to49 = "B25024_008",
                                               units_instructure_50plus = "B25024_009",
                                               contractrent_agg = "B25060_001"),       # also have rent as % of income
                                 geometry = TRUE,
                                 keep_geo_vars = TRUE,
                                 year = 2023,
                                 output = "wide") %>%
  filter(ALAND != 0) %>% # Filter tracts that are 100% water
  mutate(GEOID = as.double(GEOID))
```

## Cleaning and transforming dataset (creating percentages)
```{r}
kcacs_blocks <- kingcounty_acs_blocks %>%
  mutate(ALAND_miles = ALAND/2589988) %>% # Converting sq meters to sq miles
  mutate(
        pop_per_sqmile = total_populationE / ALAND_miles,
        race_white_perc = race_whiteE / race_basetotalE,
        race_black_perc = race_blackE / race_basetotalE,
        race_asian_perc = race_asianE / race_basetotalE,
        race_nonwhite_perc = (race_basetotalE - race_whiteE) / race_basetotalE,
        hisp_perc = hisp_hispanicE / hisp_basetotalE,
        tt_less10_perc = tt_less10E / transp_basetotalE,
        tt_10to14_perc = tt_10to14E / transp_basetotalE,
        tt_15to19_perc = tt_15to19E / transp_basetotalE,
        tt_20to24_perc = tt_20to24E / transp_basetotalE,
        tt_25to29_perc = tt_25to29E / transp_basetotalE,
        tt_30to34_perc = tt_30to34E / transp_basetotalE,
        tt_35to44_perc = tt_35to44E / transp_basetotalE,
        tt_45to59_perc = tt_45to59E / transp_basetotalE,
        tt_60plus_perc = tt_60plusE / transp_basetotalE,
        transp_mthd_public_perc = transp_mthd_publicE / transp_basetotalE,
        transp_mthd_publicbus_perc = transp_mthd_public_busE / transp_basetotalE,
        public_assistance_yes_perc = public_assistance_yesE / public_assistance_basetotalE,
        tenure_ownerocc_perc = tenure_owneroccE / tenure_basetotalE,
        tenure_renterocc_perc = tenure_renteroccE / tenure_basetotalE,
        units_instructure_1_perc = 
          (units_instructure_1detE + units_instructure_1attE) / units_instructure_basetotalE,
        units_instructure_2_perc = units_instructure_2E / units_instructure_basetotalE,
        units_instructure_3to4_perc = units_instructure_3to4E / units_instructure_basetotalE,
        units_instructure_5to9_perc = units_instructure_5to9E / units_instructure_basetotalE,
        units_instructure_10to19_perc = units_instructure_10to19E / units_instructure_basetotalE,
        units_instructure_20to49_perc = units_instructure_20to49E/ units_instructure_basetotalE,
        units_instructure_50plus_perc = units_instructure_50plusE / units_instructure_basetotalE) %>%
  select(
        STATEFP,
        COUNTYFP,
        TRACTCE,
        BLKGRPCE,
        GEOIDFQ,
        GEOID,
        NAME.x,
        ALAND,
        AWATER,
        total_populationE,
        pop_per_sqmile,
        median_incomeE,
        race_white_perc,
        race_black_perc,
        race_asian_perc,
        race_nonwhite_perc,
        hisp_perc,
        tt_less10_perc,
        tt_10to14_perc,
        tt_15to19_perc,
        tt_20to24_perc,
        tt_25to29_perc,
        tt_30to34_perc,
        tt_35to44_perc,
        tt_45to59_perc,
        tt_60plus_perc,
        transp_mthd_public_perc,
        transp_mthd_publicbus_perc,
        agg_tt_overallE,
        agg_tt_publicE,
        agg_tt_public_busE,
        public_assistance_yes_perc,
        tenure_ownerocc_perc,
        tenure_renterocc_perc,
        units_instructure_1_perc,
        units_instructure_2_perc,
        units_instructure_3to4_perc,
        units_instructure_5to9_perc,
        units_instructure_10to19_perc,
        units_instructure_20to49_perc,
        units_instructure_50plus_perc,
        contractrent_aggE,
        geometry)
```

# Assorted Visualizations

## Traffic Volume by Day/Time

```{r}
# Ridgeline Plot of Traffic by Day
# Seems that I may want to log it
library(ggridges)

congestion_temporal_ridgeline <- congestion_temporal %>%
  mutate(WEEKDAY_NAME = as.factor(case_when(
    WEEKDAY == 1 ~ "Monday",
    WEEKDAY == 2 ~ "Tuesday",
    WEEKDAY == 3 ~ "Wednesday",
    WEEKDAY == 4 ~ "Thursday",
    WEEKDAY == 5 ~ "Friday",
    WEEKDAY == 6 ~ "Saturday",
    WEEKDAY == 7 ~ "Sunday"))) %>%
  filter(TOTAL != -1,
         TOTAL != 0)

ridgeline_trafficday <- ggplot(data = congestion_temporal_ridgeline,
                               aes(x = log(TOTAL),
                                   y = WEEKDAY_NAME)) + 
  geom_density_ridges(scale = 4) +
  scale_y_discrete(expand = c(0, 0)) + 
  scale_x_continuous(expand = c(0, 0)) + 
  coord_cartesian(clip = "off") + 
  theme_ridges()

ridgeline_trafficday
```

```{r}
# Plot of Average Congestion by Hour by Day
plot_congestion_hourday <- ggplot(data = congestion_dh_longer,
                                  aes(x = HOUR,
                                      y = avg_traffic_dayhour,
                                      color = WEEKDAY_NAME,
                                      linetype = WEEKDAY_NAME)) +
  geom_line()

plot_congestion_hourday
```

## Delay Plots

```{r}
# Delay by Day
library(ggridges)

delay_day_ridgeline <- gtfs_main_withcongestion %>%
  filter(arrival_delay > -2500)

ridgeline_delayday <- ggplot(data = delay_day_ridgeline,
                               aes(x = arrival_delay,
                                   y = weekday)) + 
  geom_density_ridges(scale = 4) +
  scale_y_discrete(expand = c(0, 0)) + 
  scale_x_continuous(expand = c(0, 0)) + 
  coord_cartesian(clip = "off") + 
  theme_ridges()

ridgeline_delayday
```

```{r}
# Scatterplot of avg congestion by avg delay
delay_traffic_data <- gtfs_main_withcongestion %>%
  filter(arrival_delay > -2500) %>%
  mutate(dayhour_traffic_bins = cut(avg_traffic_dayhour, 
                                    breaks = c(-Inf, 50, 100, 150, 200, 250, 300, 350, 400, 450, 500, 550, 600, Inf),
                                    labels = c("0-50", "50-100", "100-150", "150-200", "200-250", "250-300", "300-350",
                                               "350-400", "400-450", "450-500", "500-550", "550-600", "600+")))

scatter_delaytraffic <- ggplot(data = delay_traffic_data,
                               aes(x = dayhour_traffic_bins,
                                   y = arrival_delay,
                                   color = dayhour_traffic_bins)) +
  geom_jitter() + 
  theme(legend.position = "none")
scatter_delaytraffic
```

```{r}
# Scatterplot of avg congestion by avg delay
delay_traffic_data <- gtfs_main_withcongestion %>%
  filter(arrival_delay > -2500) %>%
  mutate(dayhour_traffic_bins = cut(avg_traffic_dayhour, 
                                    breaks = c(-Inf, 50, 100, 150, 200, 250, 300, 350, 400, 450, 500, 550, 600, Inf),
                                    labels = c("0-50", "50-100", "100-150", "150-200", "200-250", "250-300", "300-350",
                                               "350-400", "400-450", "450-500", "500-550", "550-600", "600+")))

scatter_delaytraffic <- ggplot(data = delay_traffic_data,
                               aes(x = arrival_delay,
                                   y = dayhour_traffic_bins,
                                   fill = dayhour_traffic_bins)) +
  geom_density_ridges() + 
  theme(legend.position = "none") +
  xlim(-500, 500)
scatter_delaytraffic
```

```{r}
# Delay by Day/Hour
day_hour_delay <- gtfs_main_withcongestion %>%
  filter(arrival_delay > 0,
         arrival_delay < 300) %>%
  select(hr, arrival_delay, weekday) %>%
  group_by(weekday, hr) %>%
  summarize(arrival_delay = mean(arrival_delay)) %>%
  ungroup()

plot_delay_hourday <- ggplot(data = day_hour_delay,
                                  aes(x = hr,
                                      y = arrival_delay,
                                      color = weekday,
                                      linetype = weekday)) +
  geom_line()

plot_delay_hourday
```










































