---
title: "hierarchicalBART-modeling"
author: "Peter Silverstein"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
# Loading Libraries

# General Use
library(tidyverse)
library(ggplot2)
library(here)
library(arm)

# Modeling
library(stan4bart)
library(bartCause)
library(rstanarm)
```

# Loading Data

```{r}
stop_pred_std <- read_csv(here("predictor_tables","stop_pred_std.csv"), show_col_types = FALSE)
route_pred_std <- read_csv(here("predictor_tables","route_pred_std.csv"), show_col_types = FALSE)

weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
weekends <- c("Saturday", "Sunday")
peak <- c(6, 7, 8, 9, 16, 17, 18, 19) # weekdays only
non_peak <- c(1, 2, 3, 4, 5, 10, 11, 12, 13, 14, 15, 20, 21, 22, 23, 24)

data <- stop_pred_std %>%
  inner_join(route_pred_std,
             by = c("route_id" = "ROUTE_ID")) %>%
  filter(!is.na(arrival_delay)) %>%
  mutate(g_weekend = case_when(
    weekday %in% weekdays ~ 0,
    weekday %in% weekends ~ 1
  ),
  g_peak = case_when(
    weekday %in% weekdays & hr %in% peak ~ 1,
    TRUE ~ 0
  )) %>%
  dplyr::select(abs_dev, 
         rapid_ride, 
         shape_dist_traveled, 
         avg_traffic_dayhour, 
         weekday, 
         hr, 
         spatial_traffic, 
         ridership_perc,
         route_id,
         g_peak,
         g_weekend) %>%
  mutate(rapid_ride = case_when(
    rapid_ride == 1 ~ 0,
    rapid_ride == 2 ~ 1),
    weekday = case_when(
      weekday == "Monday" ~ 1,
      weekday == "Tuesday" ~ 2,
      weekday == "Wednesday" ~ 3,
      weekday == "Thursday" ~ 4,
      weekday == "Friday" ~ 5,
      weekday == "Saturday" ~ 6,
      weekday == "Sunday" ~ 7
    )) %>%
  mutate(rapid_ride = as.factor(rapid_ride),
         weekday = as.numeric(weekday),
         hr = as.numeric(hr)) %>%
  rename("g_routeid" = "route_id",
         "g_weekday" = "weekday",
         "g_hr" = "hr")
```

# Train and Test Sets/Final Variable Selection/Matrix Setup

```{r}
# JUST NEED TO SPLIT THE DATA IN HALF
set.seed(50)

# Train/Test Partition
smp_size <- floor(0.005 * nrow(data))
train_indices <- sample(seq_len(nrow(data)), size = smp_size)

train <- data[train_indices, ]
test <- data[-train_indices, ]
```

# Balance Diagnostics

```{r}
# Function to get balance diagnostics
# From https://github.com/gperrett/stan4bart-study/blob/master/get_balance.R
# Linked in Dorie et al 2022
get_balance <- function(rawdata, treat,estimand="ATT"){
  if(missing(rawdata)) stop("rawdata is required")
  if(missing(treat)) stop("treatment vector (treat) is required")
  cat("Balance diagnostics assume that the estimand is the",estimand,"\n")
  #
  #raw.dat <- data.frame(rawdata, treat = treat)
  covnames <- colnames(rawdata)
  if (is.null(covnames)){
    cat("No covariate names provided.  Generic names will be generated.")
    covnames = paste("v",c(1:ncol(rawdata)),sep="")
  }
  K <- length(covnames)
  diff.means <- matrix(NA, K, 5)
  var.t <- numeric(K)
  var.c <- numeric(K)
  std.denom <- numeric(K)
  binary <- rep(1,K)
  
  for (i in 1:K) {
    # separate means by group
    diff.means[i, 1] <- mean(rawdata[treat==1, i])
    diff.means[i, 2] <- mean(rawdata[treat==0, i])
    # separate variances by group == only used as input to calculations below
    var.t[i] <- var(rawdata[(treat == 1), i])
    var.c[i] <- var(rawdata[(treat == 0), i])
    # denominator in standardized difference calculations
    if(estimand=="ATE"){std.denom[i] <- sqrt((var.t[i]+var.c[i])/2)}
    else{
      std.denom[i] <- ifelse(estimand=="ATT",sqrt(var.t[i]),sqrt(var.c[i]))
    }
    # difference in means
    diff.means[i, 3] <- diff.means[i, 1] - diff.means[i, 2]
    # standardized difference in means (sign intact)
    diff.means[i, 4] <- abs(diff.means[i, 3]/std.denom[i])
    if(length(unique(rawdata[,covnames[i]]))>2){
      binary[i] = 0
      diff.means[i, 5] <- sqrt(var.c[i]/var.t[i])
    }
  }
  
  dimnames(diff.means) <- list(covnames, c("treat", "control", "unstd.diff",
                                           "abs.std.diff", "ratio"))
  
  
  return(diff.means)
}

# Setting up data (covariates in a matrix, treatment vector)
# Removed factor vars
X <- as.matrix(train[, c("g_hr", "g_weekday", "shape_dist_traveled", "avg_traffic_dayhour", "spatial_traffic", "ridership_perc")])
y <- train$rapid_ride

# Running the function
get_balance(rawdata = X, treat = y, estimand = "ATT")
```

# Model Setup & Fit

## ROUTEID and STOP? Hierarcichal
```{r}
model <- stan4bart(abs_dev ~ bart(.-rapid_ride-g_weekday-g_hr-g_weekend-g_peak) + rapid_ride + (1 | weekday/hr),
                   data = train,
                   treatment = rapid_ride,
                   cores = 2,
                   chains = 8,
                   iter = 2000,
                   seed = 50,
                   verbose = 2,
                   bart_args = list(n.trees = 150))
```

# Prediction/Treatment Effect Estimation

## Treatment Effects Among the Treated

Note that this is ALL conducted on the "training" set defined in a previous code block. Confusingly, the `sample = "test"` argument actually means that the function is calculating counterfactual potential outcomes, not that it is using some test data set.

```{r}
# Function to extract relevant information from the model
extract_stan4bart <- function(fit, .model) {
  
  # Extract posterior predictive distributions for test set (returns "non-treated" predictions for all units)
  samples.ppd.test <- extract(fit, type = "ppd", sample = "test")
  
  # SATT Calculation
  ## First, individual treatment effect
  samples.ite <- (train$y - samples.ppd.test) * (2 * train$rapid_ride - 1)
  
  ## Filter data to only contain treated units
  samples.ite <- samples.ite[train$rapid_ride == 1,]
  
  ## Sample average treatment effect
  samples.satt <- apply(samples.ite, 2, mean)
  satt <- mean(samples.satt)
  satt.int <- quantile(samples.satt, c(.025, .975))
  
  # Individual Conditional Average Treatment Effect (not really that applicable here but still interesting)
  samples.mu.train <- stan4bart:::extract.stan4bartFit(fit)
  samples.mu.test  <- stan4bart:::extract.stan4bartFit(fit, sample = "test")
  samples.icate <- (samples.mu.train  - samples.mu.test) * (2 * fit$frame[[fit$treatment]] - 1)
  
  # Results
  average.pred <- sate
  interval <- satt.int
  icate.pred <- apply(samples.icate, 1, mean)
  icate.pred <- icate.pred[train$rapid_ride == 1]
  
  # Subgroup Analysis 
  ## Weekday (1-7)
  g_weekday <- train$g_weekday[train$rapid_ride == 1]
  g_weekday.pred <- unlist(lapply(split(icate.pred, g_weekday), mean))
  
  ## Weekend (Yes == 1, No == 0)
  g_weekend <- train$g_weekend[train$rapid_ride == 1]
  g_weekend.pred <- unlist(lapply(split(icate.pred, g_weekend), mean))
  
  ## Peak (Yes == 1, No == 0)
  g_peak <- train$g_peak[train$rapid_ride == 1]
  g_peak.pred <- unlist(lapply(split(icate.pred, g_peak), mean))
  
  ## Route (Rapid_Ride == 1 + Route_ID)
  g_routeid <- train$g_routeid[train$rapid_ride == 1]
  g_routeid.pred <- unlist(lapply(split(icate.pred, g_routeid), mean))
  
  # Performance Metrics (NEED TO EDIT TO MATCH MY STUFF)
  out <- data.frame(
  u_bias = average.pred - average.truth,
  s_bias = (average.pred - average.truth) /sd(abs_dev),
  ci_len = (interval[2] - interval[1]) / sd(abs_dev), 
  cover = as.double(interval[1] <= average.truth && average.truth <= interval[2]),
  pehe = sqrt(mean((icate.pred - icate.truth)^2)) / sd(abs_dev),
  pegste_weekday = sqrt(mean((g_weekday.pred - g_weekday.truth)^2)) / sd(abs_dev),
  pegste_weekend = sqrt(mean((g_weekend.pred - g_weekend.truth)^2)) / sd(abs_dev),
  pegste_peak = sqrt(mean((g_peak.pred - g_peak.truth)^2)) / sd(abs_dev),
  pegste_routeid = sqrt(mean((g_routeid.pred - g_routeid.truth)^2)) / sd(abs_dev),
  model = .model, 
  tau = tau
  )
}
```

# Graphs


# Descriptive Tables as well!
# Routes, maps, thesis!!
































