---
title: "exploratory-data-viz"
author: "Peter Silverstein"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
# Loading Libraries

# General Use
library(tidyverse)
library(ggplot2)
library(here)
library(patchwork)
library(ggridges)
library(GGally)
library(rstanarm)
library(skimr)
library(keyring)
library(tidycensus)

# GIS and Mapping
library(sf)
library(tmap) 
```

# Loading Data
```{r}
data <- read_csv(here("predictor_tables","final_data.csv")) %>%
  mutate(
    g_routeid = as.factor(g_routeid), 
    stop_id = as.factor(stop_id), 
    trip_id = as.factor(trip_id), 
    rapid_ride = as.numeric(rapid_ride), 
    arrival_delay = as.numeric(arrival_delay), 
    abs_dev = as.numeric(abs_dev),
    log_abs_dev = log(abs_dev + 0.1),
    shape_dist_traveled = as.numeric(shape_dist_traveled), 
    avg_traffic_dayhour = as.numeric(avg_traffic_dayhour), 
    spatial_congestion = as.numeric(spatial_congestion),
    pop_density = as.numeric(pop_density),
    route_ridership = as.numeric(route_ridership),
    perc_white = as.numeric(perc_white),
    median_hhi = as.numeric(median_hhi),
    g_weekday = as.numeric(g_weekday), 
    g_hr = as.numeric(g_hr),
    g_peak = as.factor(g_peak),
    g_weekend = as.factor(g_weekend)
  ) %>%
  mutate(
    abs_dev = ifelse(
    abs_dev < 80000, abs_dev,
    abs(abs_dev - 86400)
    ))

```

# Descriptive Stats

## Main Descriptive Stats, by treatment group
```{r}
data_desc = data %>%
  select(
    rapid_ride, 
    log_abs_dev,
    shape_dist_traveled, 
    avg_traffic_dayhour, 
    spatial_congestion,
    pop_density,
    route_ridership,
    perc_white,
    median_hhi,
    g_weekday, 
    g_hr
  )

overall_descriptives <- skim(data_desc)
RR1_descriptives <- skim(data_desc %>% filter(rapid_ride == 1))
RR0_descriptives <- skim(data_desc %>% filter(rapid_ride == 0))

# CAN USE KABLE W/ DIGITS ARGUMENT FOR BETTER PRINTING
```

## Assessing Balance for Causal Inference
```{r}
# Function to get balance diagnostics
# From https://github.com/gperrett/stan4bart-study/blob/master/get_balance.R
# Linked in Dorie et al 2022
get_balance <- function(rawdata, treat,estimand="ATT"){
  if(missing(rawdata)) stop("rawdata is required")
  if(missing(treat)) stop("treatment vector (treat) is required")
  cat("Balance diagnostics assume that the estimand is the",estimand,"\n")
  #
  #raw.dat <- data.frame(rawdata, treat = treat)
  covnames <- colnames(rawdata)
  if (is.null(covnames)){
    cat("No covariate names provided.  Generic names will be generated.")
    covnames = paste("v",c(1:ncol(rawdata)),sep="")
  }
  K <- length(covnames)
  diff.means <- matrix(NA, K, 5)
  var.t <- numeric(K)
  var.c <- numeric(K)
  std.denom <- numeric(K)
  binary <- rep(1,K)
  
  for (i in 1:K) {
    # separate means by group
    diff.means[i, 1] <- mean(rawdata[treat==1, i])
    diff.means[i, 2] <- mean(rawdata[treat==0, i])
    # separate variances by group == only used as input to calculations below
    var.t[i] <- var(rawdata[(treat == 1), i])
    var.c[i] <- var(rawdata[(treat == 0), i])
    # denominator in standardized difference calculations
    if(estimand=="ATE"){std.denom[i] <- sqrt((var.t[i]+var.c[i])/2)}
    else{
      std.denom[i] <- ifelse(estimand=="ATT",sqrt(var.t[i]),sqrt(var.c[i]))
    }
    # difference in means
    diff.means[i, 3] <- diff.means[i, 1] - diff.means[i, 2]
    # standardized difference in means (sign intact)
    diff.means[i, 4] <- abs(diff.means[i, 3]/std.denom[i])
    if(length(unique(rawdata[,covnames[i]]))>2){
      binary[i] = 0
      diff.means[i, 5] <- sqrt(var.c[i]/var.t[i])
    }
  }
  
  dimnames(diff.means) <- list(covnames, c("treat", "control", "unstd.diff",
                                           "abs.std.diff", "ratio"))
  
  
  return(diff.means)
}

# Setting up data (covariates in a matrix, treatment vector)
# Removed factor vars
X <- as.matrix(data_desc %>% select(!rapid_ride))
y <- data_desc$rapid_ride

# Running the function
balance_table <- get_balance(rawdata = X, treat = y, estimand = "ATT")

# Need a prettier printing solution (KABLE)
balance_table
```

# Visualizations

## Maps

### Routes + Stops in the City
```{r}
# Loading Data
routes_inscope <- unique(data$g_routeid)
stops_inscope <- unique(data$stop_id)
routes_rapidride <- data %>%
  filter(rapid_ride == 1)
routes_rapidride <- unique(routes_rapidride$g_routeid)

routes_shp <- st_read(here("Predictor Data Sets",
                           "KCMetro_Transit_Lines",
                           "Transit_Routes_for_King_County_Metro___transitroute_line.shp")) %>%
  filter(ROUTE_ID %in% routes_inscope)

routes_rapidride_shp <- routes_shp %>%
  filter(ROUTE_ID %in% routes_rapidride)

stops_shp <- st_read(here("Predictor Data Sets",
                          "KCMetro_Transit_Stops",
                          "Transit_Stops_for_King_County_Metro___transitstop_point.shp"))

# Mapping
routestop_map <- tm_shape(routes_shp) + tm_lines(col = "blue") +
  tm_shape(routes_rapidride_shp) + tm_lines(col = "red") +
  tm_basemap("CartoDB.PositronNoLabels")

routestop_map
```

### Traffic in the City
```{r}
# Loading Data
congestion_spatial <- st_read(here("Predictor Data Sets",
                                   "2018_Traffic_Flow_Counts-shp",
                                   "2018_Traffic_Flow_Counts.shp")) %>%
  select(AWDT, geometry) %>%
  st_transform(crs = 2285)

# Mapping
traffic_map <- tm_shape(congestion_spatial) + tm_lines(col = "AWDT", palette = "ylOrRd", style = "sd") + 
  tm_basemap("CartoDB.PositronNoLabels")

traffic_map
```

### Faceted Maps of ACS Variables
```{r}
# Loading Data
# Using keyring package to keep my API key hidden
tidycensus_api_key <- key_get(service = "tidycensus_API", username = "my_tidycensus")
census_api_key(tidycensus_api_key)

ACSlist <- load_variables(2022, "acs5", cache = TRUE)

# Projection is NAD83(!!)
kingcounty_acs_blocks <- get_acs(state = "WA",
                                 county = "King",
                                 geography = "block group",
                                 variables = c(tot_pop = "B01003_001",
                                               transp_basetotal = "B08134_001",
                                               transp_mthd_public = "B08134_061",
                                               race_base = "B02001_001",
                                               race_white = "B02001_002",
                                               median_HHI = "B19013_001"),
                                 geometry = TRUE,
                                 keep_geo_vars = TRUE,
                                 year = 2023,
                                 output = "wide") %>%
  filter(ALAND != 0) %>% # Filter tracts that are 100% water
  mutate(GEOID = as.double(GEOID))

kcacs_blocks <- kingcounty_acs_blocks %>%
  mutate(ALAND_miles = ALAND/2589988) %>% # Converting sq meters to sq miles
  mutate(transp_mthd_public_perc = transp_mthd_publicE / transp_basetotalE,
         pop_density = tot_popE / ALAND_miles,
         white_perc = race_whiteE / race_baseE,
         median_HHI = median_HHIE) %>%
  select(tot_popE,
         pop_density,
         transp_mthd_public_perc,
         white_perc,
         median_HHI,
         geometry) %>%
  st_transform(crs = 2285)

# Mapping
popdensity_map <- tm_shape(kcacs_blocks) + tm_polygons(col = pop_density) + 
  tm_basemap("CartoDB.PositronNoLabels")

ridership_map <- tm_shape(kcacs_blocks) + tm_polygons(col = transp_mthd_public_perc) + 
  tm_basemap("CartoDB.PositronNoLabels")

whiteperc_map <- tm_shape(kcacs_blocks) + tm_polygons(col = white_perc) + 
  tm_basemap("CartoDB.PositronNoLabels")

medianHHI_map <- tm_shape(kcacs_blocks) + tm_polygons(col = median_HHI) + 
  tm_basemap("CartoDB.PositronNoLabels")

popdensity_map + ridership_map + whiteperc_map + medianHHI_map
```

## Graphs

### Traffic: Average Traffic by Day+Hour
```{r}
data_trafficdayhour <- data_desc %>%
  select(avg_traffic_dayhour,g_weekday, g_hr)

avgtraffic_plot <- ggplot(data = data_trafficdayhour,
                          aes(x = g_hr, 
                              y = avg_traffic_dayhour,
                              color = factor(g_weekday))) + 
  geom_line(aes(linetype = factor(g_weekday)))

avgtraffic_plot
```

### Graphs Per Predictor vs. Absolute Deviation
```{r}
# shape_dist_traveled = as.numeric(shape_dist_traveled)
disttraveled_plot <- ggplot(data = data,
                            aes(x = shape_dist_traveled,
                                y = log_abs_dev,
                                color = factor(rapid_ride))) +
  geom_point(size = 0.5) + 
  theme(legend.position = "none")

# avg_traffic_dayhour = as.numeric(avg_traffic_dayhour)
timetraffic_plot <- ggplot(data = data,
                            aes(x = avg_traffic_dayhour,
                                y = log_abs_dev,
                                color = factor(rapid_ride))) +
  geom_point(size = 0.5) + 
  theme(legend.position = "none")

# spatial_congestion = as.numeric(spatial_congestion)
spacetraffic_plot <- ggplot(data = data,
                            aes(x = spatial_congestion,
                                y = log_abs_dev,
                                color = factor(rapid_ride))) +
  geom_point(size = 0.5) + 
  theme(legend.position = "none")

# pop_density = as.numeric(pop_density)
popdensity_plot <- ggplot(data = data,
                            aes(x = pop_density,
                                y = log_abs_dev,
                                color = factor(rapid_ride))) +
  geom_point(size = 0.5) + 
  theme(legend.position = "none")

# route_ridership = as.numeric(route_ridership)
ridership_plot <- ggplot(data = data,
                            aes(x = route_ridership,
                                y = log_abs_dev,
                                color = factor(rapid_ride))) +
  geom_point(size = 0.5) + 
  theme(legend.position = "none")

# perc_white = as.numeric(perc_white)
percwhite_plot <- ggplot(data = data,
                            aes(x = perc_white,
                                y = log_abs_dev,
                                color = factor(rapid_ride))) +
  geom_point(size = 0.5) + 
  theme(legend.position = "none")

# median_hhi = as.numeric(median_hhi)
medianhhi_plot <- ggplot(data = data,
                            aes(x = median_hhi,
                                y = log_abs_dev,
                                color = factor(rapid_ride))) +
  geom_point(size = 0.5) + 
  theme(legend.position = "none")

# g_weekday = as.numeric(g_weekday)
weekday_plot <- ggplot(data = data,
                            aes(x = g_weekday,
                                y = log_abs_dev,
                                color = factor(rapid_ride))) +
  geom_jitter(width = 0.5, size = 0.5) + 
  theme(legend.position = "none")

# g_hr = as.numeric(g_hr)
hour_plot <- ggplot(data = data,
                            aes(x = g_hr,
                                y = log_abs_dev,
                                color = factor(rapid_ride))) +
  geom_jitter(width = 0.5, size = 0.5) + 
  theme(legend.position = "none")

(disttraveled_plot | timetraffic_plot | spacetraffic_plot) /
(popdensity_plot | ridership_plot | percwhite_plot) /
(medianhhi_plot | weekday_plot | hour_plot)
```

## Raw Estimates of Average Absolute Deviation w/ Error

### RapidRide vs. Non-RapidRide
```{r}
rapidride_stats <- data %>%
  group_by(rapid_ride) %>%
  summarize(
    mean = mean(abs_dev, na.rm = TRUE),
    se = sd(abs_dev, na.rm = TRUE) / sqrt(n()),
    n = n()
    )

ggplot(rapidride_stats, aes(x = rapid_ride, y = mean, color = factor(rapid_ride))) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  labs(
    title = "Mean Absolute Deviation by RapidRide Status",
    subtitle = "Error bars represent ± 1 standard error",
    x = "RapidRide Status",
    y = "Mean Outcome"
  ) +
  theme_minimal() +
  coord_flip()
```

### By Route
```{r}
route_stats <- data %>%
  group_by(g_routeid) %>%
  summarize(
    mean = mean(abs_dev, na.rm = TRUE),
    se = sd(abs_dev, na.rm = TRUE) / sqrt(n()),
    n = n(),
    rapid_ride = as.factor(rapid_ride)
    )

ggplot(route_stats, aes(x = g_routeid, y = mean, color = rapid_ride)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  labs(
    title = "Mean Absolute Deviation by Route",
    subtitle = "Error bars represent ± 1 standard error",
    x = "Route ID",
    y = "Mean Outcome"
  ) +
  theme_minimal() +
  coord_flip()
```












